---
title: "data_analysis_EPA"
output: html_document
date: "2023-11-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
```

```{r}
foreclosures <- read_csv("Maryland_Notices_of_Intent_to_Foreclose.csv") |> 
  clean_names() |> 
  rename(february_2022 = february_2020)

foreclosures

foreclosures_sum <- foreclosures |>
  group_by(zip) |>
  summarise(foreclosures_sum = sum(across(contains("_")))) |>
  arrange(desc(foreclosures_sum))
foreclosures_sum

write_csv(foreclosures_sum, "foreclosure_sum.csv")
```
1. From July '21- Oct'23 "the end of the pause on foreclosure notices", which zip code has had the most notices recorded?
2. How has ^^this zip code^^ changed over the timespan?
3. Using census data, what are the demographics of ^^this zip code^^?
4. During the "holiday season" (Oct-Dec), how many foreclosures occur? And where?
5. What are the differences in foreclosure noties by season (warm months vs cold months)?

https://thedailyrecord.com/2023/06/16/md-baltimore-see-increase-in-foreclosure-rate/
https://www.attomdata.com/news/market-trends/foreclosures/attom-july-2023-u-s-foreclosure-market-report/
https://www.attomdata.com/news/market-trends/foreclosures/attom-mid-year-2023-u-s-foreclosure-market-report/
https://patch.com/maryland/baltimore/baltimore-columbia-towson-md-has-one-highest-foreclosure-rates-country


