---
title: "data_analysis_EPA"
output: html_document
date: "2023-11-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
```
# Dataset: Maryland Notices of Intent to Foreclose (July 2021 - Oct. 2023)

-----------

# Questions we would like to answer throughout the course of this project:



#Q1. From July '21- Oct'23 "the end of the pause on foreclosure notices", which zip code has had the most notices recorded?

# Why is this newsworthy? 
The pandemic wasn't all that long ago, and we would like to see if its impact is still being felt financially across the state, and if so, where? Especially since the end of the moratorium on intent to foreclose notices began on July 1, 2021, when this data begins. 

Details on moratorium: https://content.govdelivery.com/accounts/MDDLLR/bulletins/2e5c0c0

# Have other news outlets reported on this? 
Other outlets have reported about the rise in foreclosures following the end of the moratorium on intent to foreclose notices, but not necessarily by zip code in Maryland. Here are the stories we found:
# This one has a paywall. I had trouble accessing it again after I exited the page. https://thedailyrecord.com/2023/06/16/md-baltimore-see-increase-in-foreclosure-rate/
https://www.attomdata.com/news/market-trends/foreclosures/attom-july-2023-u-s-foreclosure-market-report/
https://www.attomdata.com/news/market-trends/foreclosures/attom-mid-year-2023-u-s-foreclosure-market-report/
https://patch.com/maryland/baltimore/baltimore-columbia-towson-md-has-one-highest-foreclosure-rates-country



#Q2. How has ^^this zip code^^ changed over the timespan?

# Why is this newsworthy?
We want to see if the notices hit all at once, if there was a gradual incline or a steady rate of notices once the moratorium was lifted a little over two years later. The big story here would be if the notices are still steadily rising or ongoing in an area that hasn't garnered a lot of attention. 

# Have other news outlets reported on this?

Any outlets that would have touched on this would have brought it up in the stories listed above. The first and last stories above focus more on Maryland and/or Baltimore but do mention at least a pattern of rising foreclosure rates. (These are also foreclosures rather than just notices)



#Q3. Using census data, what are the demographics of ^^this zip code^^?

# Why is this newsworthy?

Who is being overlooked? Who are the people who live in a zip code that has a poor rate of notices compared to others? This is an ongoing question to ask.

# Have other news outlets reported on this?
Not that we have seen.



#Q4. During the "holiday season" (Oct-Dec), how many foreclosures occur? And where?

# Why is this newsworthy?
The holiday season is upon us. How many people found themselves without their home during these months over the past two years? What did this October look like?

# Have other news outlets reported on this?
None that we saw.



#Q5. What are the differences in foreclosure notices by season (warm months vs cold months)?

# Why is this newsworthy?
Along with the holidays, winter is also upon us. If foreclosure notices have been trending upwards, that could spell for more people to be without their home in the cold, or at least present difficulty in finding a new place if weather becomes a problem. It would also be timely for summer months when the heat is overwhelming and makes moving residency more difficult in terms of labor.

# Have other news outlets reported on this?
None that we saw.



```{r}
# Read in and clean the data
foreclosures <- read_csv("Maryland_Notices_of_Intent_to_Foreclose.csv") |> 
  clean_names() |> 
  rename(february_2022 = february_2020)
```


```{r}
# Results of code: 
foreclosures
```


```{r}
# Here we wanted to try and make a quick map to see what we were looking at other than just a table of numbers.
foreclosures_sum <- foreclosures |>
  group_by(zip) |>
  summarise(foreclosures_sum = sum(across(contains("_")))) |>
  arrange(desc(foreclosures_sum))
foreclosures_sum

write_csv(foreclosures_sum, "foreclosure_sum.csv")
```

